---
title: "Time series analysis"
author: "Sivert Selnes"
date: "\today"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("itsmr")
```

```{r}
setwd("~/Tidsrekker/TMA4285_Project_1")
dataseries <- read.table("DataEx3G14.txt")
```


### Abstract

### Introduction

### Theory
 * What is a time series
 * Calculation of ACF - and what do we use it for?
 

 * Calculation of PACF - and what do we use it for?

 * Calculation of model parameters $\phi$, $\theta$ and $\sigma^2$
 * Need the uncertainty in those as well
 
 * Calculate / arrive at model prediction at sample points
 * Calculate prediction for future values
 * Calculate uncertainty in both cases
 
 * Diagnostics: strategies for assessing model choice 
 
### Data analysis
 * The plots from 1-4 go in here, basicly most of the results of the R-code are displayed here.

### Discussion
 * Comments on prelim plots: time series, ACF, PACF
 
 * Comments on estimated paramateres and uncertainties in particular
 
 * Comments on fit/prediction of the chosen ARMA model
 
 * Finally, comment on the diagnostics, AICC
 
 
### Conclusion


## Appendix, References

