---
title: "Time series analysis"
author: "Sivert Selnes"
date: "\today"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("itsmr")
```


```{r}
setwd("~/Tidsrekker/TMA4285_Project_1")
dataseries <- read.table("DataEx3G14.txt")
```


### Abstract

### Introduction


### Theory
 * What is a time series
 * Calculation of ACF - and what do we use it for?
 * Calculation of PACF - and what do we use it for?

 * Calculation of model parameters $\phi$, $\theta$ and $\sigma^2$
 * Need the uncertainty in those as well
 
 * Calculate / arrive at model prediction at sample points
 A convenient way to produce model prediction values is to use the innovations algorithm, which simiplifies in the case of a casual ARMA-process. 
 
 Innovations algorithm:
 
 \[
 v_0 = \kappa(1,1), \\
 \theta_{n,n-k} = v_k^{-1}(\kappa(n+1,k+1)-\sum_{j=0}^{k-1}\theta_{k,k-j}\theta_{n,n-j}v_j,  0 \leq k < n, \\
 \]
 
 \[
 \hat{X} = 
 \begin{cases}
 \sum_{j=1}^n \theta_{nj}(X_{n+1-j}-\hat{X}_{n+1-j}) & 1 \leq n < m, \\
 \phi_1X_n + ... +\phi_pX_{n+1-p} + \sum_{j=1}^p \theta_{nj}(X_{n+1-j}-\hat{X}_{n+1-j}) & n \geq m.
 \end{cases}
 \]
 
 * Calculate prediction for future values
 * Calculate uncertainty in both cases
 
 * Diagnostics: strategies for assessing model choice (AICC)
 
### Data analysis
 * The plots from 1-4 go in here, basicly most of the results of the R-code are displayed here.

ARIMA
Forecast plot
AICC
### Discussion
 * Comments on prelim plots: time series, ACF, PACF
 
 * Comments on estimated paramateres and uncertainties in particular
 
 * Comments on fit/prediction of the chosen ARMA model
 
 * Finally, comment on the diagnostics, AICC
 
 
### Conclusion

<!--
What can be assumed?
- Causal time series (Mentioned in the lecture, Monday 17th Sep)

If we make other assumptions, write that down. 

-->




## Appendix, References

