---
title: "Branch Sivert"
author: "Sivert Selnes"
date: "September 11, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("itsmr")
```

```{r}
setwd("~/Tidsrekker/TMA4285_Project_1")
dataseries <- read.table("DataEx3G14.txt")
```


Preliminary investigation:

1. Exploring the data with plotting of relevant statistics

* Trends
* Seasonality
* Any apparent sharp changes in behavior
* Outliers

We want to obtain a time series with stationary residuals, i.e. residuals that are identically distributed for any time $t$. 

To begin with we try to estimate a trend in the data (if any). 

```{r}
plot.ts(dataseries)
```

The dataseries does not reveal any obvious signs of non-stationarity. There might be a seasonal component (?), with period 100. The fluctuations do not apper to grow with time, suggesting that a transformation (such as log) is not neccessary.
```{r}
M <- c("season", 5, "trend", 0) # Set data model
residuals = Resid(dataseries,M) 
test(residuals) # Test residuals for stationarity and randomness
```


We might also want to examine the data with a differencing operator applied:

```{r}
# New model
M = c("diff",1)
residuals = Resid(dataseries,M)
test(residuals)
```

