---
title: "Untitled"
author: "Gina Magnussen"
date: "September 18, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(itsmr)
# Sample mean, variance and covariance?
dataseries <- ts(read.table("DataEx3G14.txt"))
model <- arma(dataseries, p = 2, q = 1)
print(model)

# Maximum likelihood
# AICC 

```



## Model parameter estimation including uncertainty
\todo{What assumptions are / can be made?}


* Have a stationary process / series of residuals (trend and seasonality removed)
* Fit an ARMA model to the residuals 
* Estimate \phi, \theta and \sigma^2 when p and q are known.

\begin{itemize}
  \item Weakly / strictly stationary
  \item Causal
  \item Existence and uniqueness
  \item ACF / PACF
  \item AICC criterion
\end{itemize}


### Notes

* ARMA models is a class of linear time series models
* ARMA is a parametric family 
* MSE of forecasts will depend on errors from estimation of parameters of the model (will be larger for higher-order models)



### Questions

* Goodness of fit test: Error terms/residuals white noise or not? QQ-plot

* How much theory should be included, how much should be justified?





